---
title: ''
output:
  html_document:
    toc: no
    toc_float: yes
    collapsed: no
    number_sections: no
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
---

## NYC Public Schools Matching {.tabset}

### Overview

Project in Progress:

Matched NYC schools on SC vs. no SC option


A core principle of the Individuals with Disabilities in Education Act (IDEA) is that students with special needs are placed in the Least Restrictive Environment. The LRE is the environment where students with special needs are educated alongside their peers without special needs to the maximum extent possible. The Most Restrictive Environment within a general education school is a self-contained classroom, where all students have special needs. Opponents of self-contained classrooms promote “inclusion” models, where all students (regardless of disability status) are educated in the same classroom, and that various disability mandates are accomdated through support services. These opponents of segregated special education services argue that self-contained classrooms are a form of disability segregation, and that exclusion fosters lower standards for special education students. A less explored impact of inclusion/exclusion is how special education services affect the entire school culture. When students with disabilities learn alongside students without disabilities, all students in the school might feel a greater sense of belonging or connection with the school. In New York City, some public schools have a self-contained option for students with special needs, and other public schools do not have an option for a self-contained classroom. Matching schools on a variety of demographic covariates, I explore the effect of the self-contained class option on Attendance Rates.


### Data
Data sample

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
######
# STEP 1
#load 2017-2018 third grade data
#######
library(readr)
library(tidyverse)
setwd("/cloud/project/project_scripts")
third_grade_data <- read.csv("../data/third_grade_data_2017.csv")

#boroughs
third_grade_data <- third_grade_data %>%
  mutate(borough = case_when(str_detect(`DBN`, "X") ~"Bronx",
                             str_detect(`DBN`, "Q") ~ "Queens",
                             str_detect(`DBN`, "K") ~"Brooklyn",
                             str_detect(`DBN`, "M") ~ "Manhattan",
                             str_detect(`DBN`, "R") ~"Staten Island"))
table(third_grade_data$borough)



######
# STEP 2
#mutate into whether or not it has a self-contained option
#######
## check distinct
#768 unique schools
length(unique(third_grade_data$DBN))
third_grade_data <- third_grade_data %>%
  mutate(GT = case_when(is.na(G.T) ~ 0),
         GenEd = case_when(is.na(Gen.Ed) ~ 0),
         GT = case_when(is.na(G.T) ~ 0),
         ICT_2 = case_when(is.na(ICT) ~ 0),
         SC_12.1 = case_when(is.na(SC.12.1) ~ 0),
         SC_12.1.1 = case_when(is.na(SC.12.1.1) ~ 0),
         SC_15.1 = case_when(is.na(SC.15.1) ~ 0),
         SC_6.1.1 = case_when(is.na(SC.6.1.1) ~ 0),
         SC_8.1.1 = case_when(is.na(SC.8.1.1) ~ 0))
                        
third_grade_data <- third_grade_data %>%
  mutate(GT= ifelse(is.na(GT), 1, 0),
         GenEd= ifelse(is.na(GenEd), 1, 0),
         ICT_2= ifelse(is.na(ICT_2), 1, 0),
         SC_12.1= ifelse(is.na(SC_12.1), 1, 0),
         SC_12.1.1= ifelse(is.na(SC_12.1.1), 1, 0),
         SC_15.1= ifelse(is.na(SC_15.1), 1, 0),
         SC_6.1.1= ifelse(is.na(SC_6.1.1), 1, 0),
         SC_8.1.1= ifelse(is.na(SC_8.1.1), 1, 0))

third_grade_data <- third_grade_data %>%
  mutate(self_contained = ifelse(SC.12.1 == 1 | SC_12.1.1 == 1 |
                                   SC.15.1 ==1 | SC_6.1.1 ==1 | SC_8.1.1==1 , 
         1, 0))
third_grade_data <- third_grade_data %>%
  mutate(self_contained_option = ifelse(is.na(self_contained), 0, 1))
third_grade_data <- third_grade_data %>%
  mutate(gifted_talented_option = ifelse(GT == 1, 
                                 1, 0))
third_grade_data <- third_grade_data %>%
  mutate(ICT_option = ifelse(ICT_2 == 1, 
                                  1, 0))
third_grade_data <- third_grade_data %>%
  mutate(self_contained_and_GT = ifelse(gifted_talented_option == 1 &
                                          self_contained_option == 1, 
                             1, 0))
third_grade_data <- third_grade_data %>%
  mutate(self_contained_and_GT_and_ICT = ifelse(gifted_talented_option == 1 &
                                          self_contained_option == 1 &
                                            ICT_option == 1, 
                                        1, 0))
third_grade_data <- third_grade_data %>%
  mutate(only_gen_ed = ifelse(gifted_talented_option == 0 &
                                                  self_contained_option == 0 &
                                                  ICT_option == 0 &
                                GenEd == 1, 
                                                1, 0))
write.csv(third_grade_data, '/cloud/project/data/third_grade_data_cleaned.csv')
######
# STEP 3
#find numbers/sample size
#######
##527 schools have a self contained option, 221 don't
third_grade_data %>%
  group_by(self_contained_option) %>%
  summarise(length(unique(DBN)))
## only 84 schools have a GT option
third_grade_data %>%
  group_by(gifted_talented_option) %>%
  summarise(length(unique(DBN)))
## 696 have a GenEd option for third grade
## the ones that don't have only ICT
third_grade_data %>%
  group_by(only_gen_ed) %>%
  summarise(length(unique(DBN)))

working_data <- third_grade_data %>%
  select(DBN, Percent_Attendance, Demographic.Variable, X..Poverty, borough,
         self_contained_option, gifted_talented_option, X..Male, X..Black,
         X..Female, X..White, Total.Enrollment, Economic.Need.Index, Percent_Chronically_Absent,
         X..Students.with.Disabilities)

## for each borough
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = Economic.Need.Index)) +
  facet_wrap(~borough)
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = X..Black)) +
  facet_wrap(~borough)
# for whether or not it has a self-contained option
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = X..Black)) +
  facet_wrap(~self_contained_option)
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = X..Black)) +
  facet_wrap(~gifted_talented_option)
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = X..Poverty)) +
  facet_wrap(~self_contained_option)
ggplot(data = third_grade_data) +
  geom_histogram(mapping = aes(x = X..Poverty)) +
  facet_wrap(~gifted_talented_option)

######
# STEP 3
#percent attendance for those with a self-contained option
#######
working_data <- working_data %>%
  filter(Percent_Attendance != 's') %>%
  filter(Percent_Chronically_Absent != 's')
working_data$Percent_Attendance <- as.numeric(working_data$Percent_Attendance)
attendance_compare <- working_data %>%
  group_by(self_contained_option, Demographic.Variable) %>%
  summarise(mean_attendance = mean(Percent_Attendance))
working_data %>%
  group_by(gifted_talented_option) %>%
  summarise(mean_attendance = mean(Percent_Attendance))
write.csv(working_data, file = '/cloud/project/data/third_grade_data_cleaned.csv')





```


### Associative Analysis


Associative trends

```{r, eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(readr)
library(tidyverse)
working_data <- read.csv('/cloud/project/data/third_grade_data_cleaned.csv')
## is the average attendance score among SC different than
## average attendance score among non SC
# install.packages("ggthemes")
library(ggthemes)
working_data_graph <- working_data %>%
  filter(Demographic.Variable %in% c("SWD", "All Students",
                                     "White", "Black",
                                     "Male", "Female",
                                     "Hispanic", "Poverty",
                                     "Not SWD")) %>%
  group_by(self_contained_option, Demographic.Variable) %>%
  summarise(mean = mean(Percent_Attendance),
            mean_absent = mean(Percent_Chronically_Absent))
working_data_graph <- working_data_graph %>%
  mutate(self_contained_option = ifelse(self_contained_option == 1, "Self-Contained Option",
                                        "No Self-Contained Option"))
ggplot(data = working_data_graph, aes(x = reorder(Demographic.Variable, -mean_absent),
                                y = mean_absent,
                                fill = "red")) +
         geom_bar(sta = "identity") +
  facet_wrap(~self_contained_option) +
  xlab ("Demographics") +
  ylab("Percent Chronically Absent") +
  theme_tufte() +
  geom_text(aes(label = round(mean_absent,2), vjust = -.3))

working_data %>%
  group_by(self_contained_option) %>%
  summarise(mean = mean(X..Students.with.Disabilities))
            
```


Differences in covariates

A Welch Two sample T. test confirms that there is a statistically significant difference in the rates of attendance and absenteeism. Attendance rates for the two groups differed significantly according to Welch’s t-test, t(394.4) = 6.84, p <.001. On average, schools with a self contained option have average attendance rates among SPED students of 91.4 %, and schools without a self contained option have average attendance rates among SPED students of 93.0%. A more alarming difference is the difference in chronic absenteeism rates, where the two groups again differed significantly according to Welch’s t-test, t-test, t(351.63) = -5.8884, p <.001. We are 95% confident that the difference between the Chronic Absenteeism rates between school types are from 11 to 5.5%.



```{r, eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(pacman)
#small sample matching
library(readr)
library(pacman)
p_load(MatchIt, dplyr, survey, tableone, twang, ipw, ggplot2)
working_data <- read.csv('/cloud/project/data/third_grade_data_cleaned.csv')
working_data <- working_data %>%
  filter(Demographic.Variable == "SWD")
working_data %>%
  group_by(self_contained_option) %>%
  summarise(mean_attendance = mean(Percent_Attendance),
            mean_chronic_absent = mean(Percent_Chronically_Absent))
p_load(MatchIt, dplyr, survey, tableone, twang, ipw, ggplot2)
# find the diffrences on the covariates
school_covariates <- c('X..Poverty', 'X..Black', 'X..Male', 'Economic.Need.Index',
                       'X..White', 'X..Female', 'Total.Enrollment', 'X..Students.with.Disabilities')
working_data %>%
  group_by(self_contained_option) %>%
  select(one_of(school_covariates)) %>%
  summarise_all(funs(mean(., na.rm=T)))
# do a t.test to find if there are statistically different differences
# lapply(school_covariates, function(v) {
#   t.test(working_data[, v] ~ working_data[ ,'self_contained_option'])
# })


```

### Matching

P(X = 1|X) is the probability that the school does NOT have a SC option
P(1): Probability of Treatment (no SC option)
P(0): Probability of Control (SC option)

It might seem backwards that the Treatment here is “no self-contained” option, when SC option is usually viewed as a “service”, but since the hypothesis is that schools with no SC option have higher attendance rates, and since 203 treatment schools that do not have a self-contained option, and 534 control schools do have a self-contained option, we define no SC as the Treatment.


```{r, eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
### 
##PART 1 MATCHING with Propensity Score
###
library(readr)
library(pacman)
p_load(MatchIt, dplyr, survey, tableone, twang, ipw, ggplot2)
working_data <- read.csv('/cloud/project/data/third_grade_data_cleaned.csv')
working_data <- working_data %>%
  filter(Demographic.Variable == "SWD")


working_data %>%
  group_by(self_contained_option) %>%
  summarise(mean_attendance = mean(Percent_Attendance),
            mean_chronic_absent = mean(Percent_Chronically_Absent))
with(working_data, t.test(Percent_Attendance ~ self_contained_option))
with(working_data, t.test(Percent_Chronically_Absent ~ self_contained_option))
table(working_data$self_contained_option)
## the t test is statistically signifiant, but that
# is done without any matching
# find the diffrences on the covariates
school_covariates <- c('X..Poverty', 'X..Black', 'X..Male', 'Economic.Need.Index',
                       'X..White', 'X..Female', 'Total.Enrollment', 'X..Students.with.Disabilities')
working_data %>%
  group_by(self_contained_option) %>%
  select(one_of(school_covariates)) %>%
  summarise_all(funs(mean(., na.rm=T)))
# do a t.test to find if there are statistically different differences

lapply(school_covariates, function(v) {
  t.test(working_data[, v] ~ working_data[ ,'self_contained_option'])
})
# set seed
set.seed(1731)
## issue with the matching, so what if we switch treatment and control
working_data$treatment_indicator <- ifelse(working_data$self_contained_option == 1, 0, 1)
working_data$new_outcome_labelled <- ifelse(working_data$treatment_indicator == 1, "no SC", "SC")
library(MatchIt)

school_nearest <- matchit(formula = treatment_indicator ~ Economic.Need.Index +
                            X..Black + X..Male + X..Poverty + X..Female + X..White #+ Total.Enrollment
                          + X..Students.with.Disabilities
                          , data = working_data,
                          method = "nearest",
                          family = "binomial",
                          caliper = 0.25)
summary(school_nearest)
table(working_data$treatment_indicator)
plot(school_nearest)


#create the matched set
nearest_matched <- match.data(school_nearest)

#350 schools were matched
dim(nearest_matched)

## now look at the means of the covariates
#matching was successful because the poverty rates are around .72 now together
nearest_matched %>%
  group_by(self_contained_option) %>%
  select(X..Poverty) %>%
  summarise_all(funs(mean))

## also can conduct a t test to assess the matches
with(nearest_matched, t.test(Percent_Attendance ~self_contained_option))

### ## estimating treatment effects with matched data
model <- lm(Percent_Attendance ~ treatment_indicator, data = nearest_matched)
summary(model)
table(nearest_matched$treatment_indicator)
# if self contained is 1 (treatment), you have SC option, you have higher attenance
## treatment effects without matched data
model <- lm(Percent_Attendance ~ treatment_indicator, data = working_data)
summary(model)

```




### IPTW

To obtain an unbiased estimate of the ATE, we use IPTW weights to create a synthetic sample where treatment assignment is independent of the measured baseline covariate. Each subject's weight is equal to the inverse of the probability of receiving the treatment that the subject received. IPTW helps to create a weighted sample in which the distribution of covariates is the same between Treatment and Control subjects. IPTW allows us to obtain unbiased estimates of the ATE (Austin & Stuart 2015). The IPTW yields a small standardized mean differences (less than a standard deviation), in see Table X in Appendix, and using these weights in the outcome model, we arrive at a significant effect of the treatment, meaning that attendance is greater in schools with inclusion, controlling for all of the covariates of race, gender, and percent of students with special needs.

```{r, eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}


## IPTW
#add the propensity scores
working_data$ps <- school_nearest$distance
#estimate the effect of SC option using IPTW
#create IPTW weights - 0 is no SC, which is the treatment
working_data$iptw <- ifelse(working_data$new_outcome_labelled == 'no SC', 1/(working_data$ps),
                            1/(1-working_data$ps))
#stabilized weights
working_data$stable.iptw <- ifelse(working_data$new_outcome_labelled == 'no SC',
                                   (mean(working_data$ps[working_data$new_outcome_labelled == 'no SC'])/working_data$ps),
                                   (mean(1-working_data$ps[working_data$new_outcome_labelled == 'SC'])/(1-working_data$ps)))
working_data_nomiss<- working_data %>%
  select(Percent_Attendance, treatment_indicator, X..Poverty, X..Male, X..Black,
         Economic.Need.Index, X..Female, X..White, X..Students.with.Disabilities, ps, iptw, stable.iptw,
         treatment_indicator)
#weighted data - create a weighted version of the data 
working_data_weighted <- svydesign(ids = ~1, data = working_data_nomiss, weights = working_data_nomiss$iptw)
#check the balance
SC_iptw_table <- svyCreateTableOne(vars = school_covariates, strata = "treatment_indicator", data = working_data_weighted,
                                   test = F)
SC_iptw_table
print(SC_iptw_table, smd=T)
boxplot(working_data_nomiss$X..Black ~ working_data_nomiss$treatment_indicator)
### check the covariate distribution for all the weights

for(i in 1:nrow(working_data_nomiss)){
  working_data_nomiss$X..Black[i] <- ifelse(working_data_nomiss$treatment_indicator == 1,
                                            (working_data_nomiss$X..Black[i]*(mean(working_data_nomiss$ps[working_data_nomiss$treatment_indicator==1])))/working_data_nomiss$ps[i],
                                            (mean(1-(working_data_nomiss$ps[working_data_nomiss$treatment_indicator ==0]))*working_data_nomiss$X..Black[i])/(1-working_data_nomiss$ps[i]))
}
# plot the weights
ipwplot(working_data_nomiss$ps, logscale = F,
        main = "propensity scores")
#lastly, test if distributions of covaraites are similar/diffeernt before or after weighting
ks.test(working_data_nomiss$X..Black[working_data_nomiss$treatment_indicator == 1],
        working_data_nomiss$X..Black[working_data_nomiss$treatment_indicator == 0])
#estimate the ate
mod_out_iptw <- lm(Percent_Attendance ~ treatment_indicator, weights = working_data_nomiss$iptw, 
                   data = working_data_nomiss)
##sig ate here
summary(mod_out_iptw)
```


### Subclassification and Heterogeneous Treatment Effects

To examine the heterogeneous effect of an SC option on attendance, I construct a propensity score stratum based on estimated propensity scores. Subclassification groups units into subclasses in which the distribution of covaraites are the same, and the propensity score is balanced. Five or six subclasses is the rule of thumb (Cochran 1968; Rosenbaum and Rubin 1984), but as n increases, more subclasses are better. Using 5 subclasses, we see that there is a signfiiant effect for Subclass 1, which is, the subclass with the highest proportion of Black students and highest Mean Economic Need. In other words, the effect of inclusion in increasing attendance is most strong for the neediest schools.
```{r, eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}

mod2 <- matchit(formula = treatment_indicator ~ Economic.Need.Index +
                   X..Black + X..Male + X..Poverty + X..White + X..Female #Total.Enrollment
               + X..Students.with.Disabilities
                , data = working_data_nomiss,
                method = "subclass", subclass = 5)
wd_nomiss2 <- data.frame(cbind(working_data_nomiss, match.data(mod2)[,c("distance", "subclass")]))                
head(wd_nomiss2)
table(working_data_nomiss$treatment_indicator)
## check out subclasses
wd_nomiss2$subclass <- as.factor(wd_nomiss2$subclass)
wd_nomiss2$X..Poverty <- as.factor(wd_nomiss2$X..Poverty)
wd_nomiss2 %>%
  group_by(subclass, treatment_indicator) %>%
  summarise(mean_ps = mean(ps),
            mean_economic_need = mean(Economic.Need.Index),
            mean_white = mean(X..White),
            mean_disability = mean(X..Students.with.Disabilities))
## so, all students in subclass 3 have similar propensity scores, etc.
table(wd_nomiss2$treatment_indicator)
dat <- wd_nomiss2[,c("distance", "treatment_indicator", "subclass")]
dat$Observations <- rep("NoSC", length(wd_nomiss2$treatment_indicator))
dat$Observations[dat$treatment_indicator == 0] <- "SC"
dat$ymax <- 1
quant <- quantile(wd_nomiss2$distance, probs = seq(0,1,1/5))
q <- data.frame(id = names(quant), values = unname(quant), stringsAsFactors = FALSE)
pp <- ggplot(data = dat, aes(x = distance, group = Observations))
pp + geom_density(aes(x = distance, linetype = Observations), size = 0.75, data = dat)+
  xlab("Propensity Score Logit") +
  ylab("Density") +
  geom_vline(xintercept = quant[(2:5)], linetype = "dashed") +
  theme_bw() +
  theme(legend.position = "bottom")


##estimate the ATE

mod_out_sub <- lm(Percent_Attendance ~ treatment_indicator +factor(subclass) + factor(subclass) *treatment_indicator -1, 
                  data = wd_nomiss2)
summary(mod_out_sub)



```





